Fusing Filters with Integer Linear Programming

### 2014-05-13 2014betterfusionforfilters/section/04-ILP.tex

\begin{figure}[ht!]
\begin{center}
\includegraphics[scale=0.5]{figures/ex2-normalizeInc.pdf}
\end{center}
\caption{Possible clusterings for \texttt{normalizeInc}}
\end{figure}

@@

\begin{figure}[H]
\begin{code}
 normalizeInc :: Array Int -> Array Int
 normalizeInc xs
  = let incs = map  (+1)    us      (A1) (B2)
        sum  = fold (+) 0   us      (A1) (B1)
        norm = map  (/ sum) incs    (A2) (B2)
    in  norm
\end{code}
\end{figure}

For @normalizeInc@, the first approach is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @A1@), then use a second loop to compute @norm@ (indicated by @A2@). The other approach is to compute @sum@ first (@B1@), and fuse the computation of @incs@ and @norm@ into the second loop (@B2@). For this specific function the second approach seems preferable because it avoids creating the intermediate array @incs@, but for large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.

Our objective, then, is not to simply reduce the number of loops as in~\cite{kennedy1993typedfusion}, but to reduce overall memory bandwidth, manifest arrays, and loops.
The objective of reducing manifest arrays, or maximising array contraction, has been discussed in~\cite{gao1993collective, darte2002contraction}.

In order to deal with the combinatorial explosion of potential clusterings, we describe an integer linear programming formulation.
Integer linear programming problems are defined as a set of variables, an objective linear function and a set of linear constraints.
The integer linear solver finds an assignment to the variables that minimises the objective function, while satisfying all constraints.
This problem is NP-hard, but existing ILP solvers are able to adequately solve practical problems.
Most combinator programs are small enough for this approach to be feasible, but for larger problems an approximation could be found, within a certain threshhold of the objective function.

>>>>

\begin{figure}[ht!]
\begin{center}
\includegraphics[scale=0.5]{figures/ex2-normalizeInc.pdf}
\end{center}
\caption{Possible clusterings for \texttt{normalizeInc}}
\label{f:normInc}
\end{figure}

@@

\begin{figure}[H]
\begin{code}
 normalizeInc :: Array Int -> Array Int
 normalizeInc xs
  = let incs = map  (+1)    us      (A2) (B1)
        sum  = fold (+) 0   us      (A1) (B1)
        norm = map  (/ sum) incs    (A2) (B2)
    in  norm
\end{code}
% swapped A and B, so that the first option matches with the left graph, the second with the right graph in the diagram
For @normalizeInc@, the first option is to compute @sum@ first (@A1@), and fuse the computation of @incs@ and @norm@ into the second loop (@A2@), as shown in the first graph in Figure~\ref{f:normInc}. The other option is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @B1@), then use a second loop to compute @norm@ (indicated by @B2@). For this specific function, it is fairly obvious that Option 1 is preferable because it avoids creating the intermediate array @incs@. For large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.

The @normalizeInc@ example demonstrates why it is, in general, not sufficient, to simply minimize the number of loops in a program: we have to assign weights to different types of loops and take those into account when choosing a possible clustering. Ordered typed fusion \cite{kennedy1993typedfusion} addresses this problem in the context of fusing imperative parallel and sequential loops, but also only in the absence of filter-like operations. Our problem can be directly mapped to an integer linear program with some integral, boolean and real variables, an objective function to minimise, and a set of contraints the variables must conform to. While finding a variable assignment that satisfies the constraints and is the minimal objective function is NP-hard, we can build on exisiting work here and use the best ILP-solver available, which tend to be sufficient for realistic problem sizes.  For larger problems, we can find an approximate answer, within say $10\%$ of the optimal answer, which still results in a very efficient, if not necessarily optimal program.

%For @normalizeInc@, the first approach is to fuse the computation of @incs@ and @sum@ into a single loop (indicated by @A1@), then use a second loop to compute @norm@ (indicated by @A2@). The other approach is to compute @sum@ first (@B1@), and fuse the computation of @incs@ and @norm@ into the second loop (@B2@). For this specific function the second approach seems preferable because it avoids creating the intermediate array @incs@, but for large programs with many operators the optimal way to cluster operators into loops can be entirely non-obvious.
